---
title: "Readme for Getting and Cleaning Data Course Project"
author: "Joris Schut"
date: "Friday, February 06, 2015"
output:
  html_document:
    keep_md: yes
---

## Project Description
In this project, raw data from the UCI machine learning repository was transformed into a clean data file that follows the [principles of tidy data](http://vita.had.co.nz/papers/tidy-data.pdf).

To do this a R script called run_analysis.R was created that does the following:
 0. Checks if the data is present and downloads it if this is not the case.
 1. Merges the training and the test sets to create one data set.
 2. Extracts only the measurements on the mean and standard deviation for 
    each measurement. 
 3. Uses descriptive activity names to name the activities in the data set
 4. Appropriately labels the data set with descriptive variable names. 
 5. From the data set in step 4, creates a second, independent tidy data 
    set with the average of each variable for each activity and each subject.

## Data gathering
### Source data
The UCI HAR Data set can be downloaded from [this web page](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones)
    or directly from [here](https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip)

###Collection of the raw data [1]
The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained data set has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain.

A video of the experiment including an example of the 6 recorded activities with one of the participants can be seen in the following link: [Web Link](https://www.youtube.com/watch?v=XOEN9W05_4A)

###Attribute information [1]
For each record in the data set it is provided: 
- Triaxial acceleration from the accelerometer (total acceleration) and the estimated body acceleration. 
- Triaxial Angular velocity from the gyroscope. 
- A 561-feature vector with time and frequency domain variables. 
- Its activity label. 
- An identifier of the subject who carried out the experiment.

## Data processing
Data processing took place using the run_analysis script. This script takes the raw data and transforms it into a single tidy data set. In this section the code of this script is described.

```{r, results='hide', warning=FALSE, message=FALSE}

#Step 0: download the file (if not already done) and unzip it

#Check if the zipfile containing the raw data is present and download it if this is not the case.
if (!file.exists("phonedata.zip")) {
  url  = "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
  dest = "phonedata.zip"
  meth = "internal"
  quit = TRUE
  mode = "wb"
  download.file(url, dest, meth, quit, mode)
  #Works on tested operating system (Windows 7). Please change values if needed.
  
  #Unzip the zipfile into the working directory.
  unzip("phonedata.zip")
} 

# Load the required libraries
library(dplyr)
library(stringr)
library(tidyr)

print("step 0 complete")

### Step 1: read in training data sets, apply initial column names and combine them into a combined variable

#Read in the raw data files (training sets)
features <- read.table("./UCI HAR Dataset/features.txt", stringsAsFactors = FALSE)
xtrn <- read.table("./UCI HAR Dataset/train/X_train.txt", nrows = 7352)
ytrn <- read.table("./UCI HAR Dataset/train/y_train.txt")
strn <- read.table("./UCI HAR Dataset/train/subject_train.txt")

#Read in the column names (training sets)
colnames(xtrn) <- features$V2
colnames(ytrn) <- "Activity"
colnames(strn) <- "Subject"

#Bind the differnt raw data files (training sets)
train <- cbind(strn, ytrn, xtrn)

# Read in test data sets, apply initial column names and combine them into a combined variable

#Read in the raw data files (test sets)
xtest <- read.table("./UCI HAR Dataset/test/X_test.txt", nrows = 2947)
ytest <- read.table("./UCI HAR Dataset/test/y_test.txt")
stest <- read.table("./UCI HAR Dataset/test/subject_test.txt")

#Read in the column names (test sets)
colnames(xtest) <- features$V2
colnames(ytest) <- "Activity"
colnames(stest) <- "Subject"

#Bind the differnt raw data files (test sets)
test <- cbind(stest, ytest, xtest)

# Combine both the train and test data sets into a single data frame (df)
df <- rbind(test, train)

print("step 1 complete")

### Step 2: filter columns to just the activity, subject, and measurement columns with mean and std (standard deviation) and call data frame skinny

keepnames <- grepl("mean|std|Subject|Activity", colnames(df)) & 
             !grepl("meanFreq", colnames(df))

skinny <- df[, keepnames]

print("step 2 complete")

### Step 3: change Activities from numeric to names 

# Load in activity names file and add appropriate column names prior to join.
activity_names <- read.table("./UCI HAR Dataset/activity_labels.txt", stringsAsFactors = FALSE)
colnames(activity_names) <- c("Activity", "ActDesc")

# Join activity names with skinny by Activity in each data frame.
skinny <- tbl_df(skinny)
skinny <- inner_join(skinny, activity_names)

# Reorder columns to move new activity names to first column and remove and rename description as activity
skinny <- skinny[,c(69, 1:68)]

colnames(skinny)[1] <- "Activity"

print("step 3 complete")
### Step 4: rename variable names in skinny to make more sense and separate by "-"

#Store the column names in a temporary variable
pracnames <- names(skinny)

#Change the names in the temporary variable
pracnames <- str_replace_all(pracnames, "Acc", "-acceleration-")
pracnames <- str_replace_all(pracnames, "Gyro", "-gyroscope-")
pracnames <- str_replace_all(pracnames, "Mag", "-magnitude")
pracnames <- str_replace_all(pracnames, "\\(\\)", "")
pracnames <- str_replace_all(pracnames, "^t", "time-")
pracnames <- str_replace_all(pracnames, "^f", "frequency-")
pracnames <- str_replace_all(pracnames, "tBody", "time-body-")
pracnames <- str_replace_all(pracnames, "BodyBody", "body")
pracnames <- str_replace_all(pracnames, "--", "-")
#Set the names to lower case
pracnames <- tolower(pracnames)

#Replace the names in skinny with the ones in the temporary variable.
names(skinny) <- pracnames

print("step 4 complete")

### Step 5: use tidyr to gather measurements into tidy data (1 line per observation)

#Take the skinny variable and collapse columns into key-value pairs.
ttidy <- skinny %>% gather(sensor, Value, 4:69)

# Aggregate the data, order the columns and order the rows.
ttidy <- aggregate(Value ~ sensor + subject + activity, data=ttidy,
                    mean, na.rm=TRUE)%>%
        select(subject, activity, sensor, Value)%>%
        group_by(subject, activity, sensor)

#Make all the column names start with a capital letter to look more consistent.
colnames(ttidy) <- c("Subject", "Activity", "Sensor", "Value")

# Write out tidy data set to a .txt file.
write.table(ttidy, "tidy_data.txt", row.name = FALSE)

print("step 5 complete")
```

#Tidy data file
The result of the data processing were saved in a separate text file (tidy_data.txt). This file follows the [principles of tidy data](http://vita.had.co.nz/papers/tidy-data.pdf). The result consists of 4 variables and 11880 rows. The variables used in the tidy file are the following:

```{r, echo=FALSE}
tidy <- read.table("tidy_data.txt", header=TRUE)
print(names(tidy))
```

All variables are detailed described in the Code Book.md. Please refer to this document for a more detailed explanation of the variables.

##Sources
1. [UCI Machine learning repostiory](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones)
