---
title: "Assignment 1 simulation"
author: "Joris Schut"
date: "Saturday, February 28, 2015"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

##Introduction
This document was made in the context of the statistical inference MOOC by Johns Hopkins University as part of the Data Science specialization on Coursera. In this project you will investigate the exponential distribution in R and compare it with the Central Limit Theorem.

##1. Show the sample mean and compare it to the theoretical mean of the distribution.

First, the parameters for the sample size (n; 40), lambda (0.2) and the number of experiments (runs; 1000) was set (values for the parameters were given in the assignment details). In order to create a reproducible analysis, the seed was set to 111.

```{r, echo=FALSE, fig.width=5, fig.height=3}

#Set parameters
set.seed(111)
n = 40
lambda = 0.2
runs = 1000

samples <- replicate(runs,rexp(40,lambda))
sampleMeans <- apply(samples,2,mean)
sampleVars <- apply(samples,2,var)

```

Based on the characteristics of the exponential distribution, the theoretical mean (1/lambda) was calculated for 1000 experiments of  a distribution of 40 exponentials.

```{r, echo=FALSE}
meanTheory <- 1/lambda
```

The distribution of means was plotted in a histogram together with the theoretical (blue line) and experimental (red line) averages.

```{r, echo=FALSE}
hist(sampleMeans)
abline(v=mean(sampleMeans), col="red", lwd=4)
abline(v=meanTheory, col="blue", lwd=2)
```


##2. Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution
Based on the characteristics of the exponential distribution, the theoretical variance ((1/lambda)^2 was calculated for 1000 experiments of a distribution of 40 exponentials.

```{r,echo=FALSE}
meanVariance <- (1/lambda)^2
```

The distribution of means was plotted in a histogram together with the theoretical (blue line) and experimental (red line) averages.

```{r, echo=FALSE}
hist(sampleVars)
abline(v=mean(sampleVars), col="red", lwd=1)
abline(v=meanVariance, col="blue", lwd=1)
```


##3 Show that the distribution is approximately normal

The Central Limit Theorem tells us that the sampling distribution of the mean of a population will be normally distributed, even if the population itself is not.

```{r, echo=FALSE, fig.width=5, fig.height=3}

p <- ggplot() +
geom_histogram(aes(x=sampleMeans,y=..density..),binwidth=0.1) +
labs(title="Distribution of Sample Means") +
geom_vline(aes(xintercept=meanTheory,colour="green"),size=1) +
geom_vline(aes(xintercept=mean(sampleMeans),colour="blue"),size=1) +
stat_function(aes(x=sampleMeans),fun=dnorm,arg=list(mean=meanTheory,sd=sqrt(varTheory/40)),colour="red",size=1)
print(p)

```

As the normal approximation fits pretty good with the histogram, normality of the distribution of the mean can be assumed.

\newpage

##Annex
Code used in this analysis:
```{r, fig.keep='none',  results='hide', warning=FALSE, message=FALSE}

#Set parameters
set.seed(111)
n = 40
lambda = 0.2
runs = 1000

#Determine the experimental mean
expvalues <- rexp(n, lambda)
expmean1 <- mean(expvalues)

hist(expvalues, main="Histogram of the random sample", xlab="Sample values")

#Calcualte the theoretical mean[1]
theomean1 <- 1/lambda
print(paste("Experimental mean (1 experiment): ", expmean1, "Theoretical mean: ", theomean1))

#Determine the absolute difference between the theoretical and experimental mean
diffmean1 <- abs(theomean1 - expmean1)
print(paste("Differnce in means: ", diffmean1))

#Determine the experiental variance
expvar1 <- var(expvalues)

#Calculate the theoretical variance[1]
theovar1 <- 1/(lambda^2)
print(paste("Experimental variance: ", expvar1, "Theoretical variance: ", theovar1))

#Determine the absolute difference between the theoretical and experimental variance
diffvar1 <- abs(theovar1 - expvar1)
print(paste("Differnce invariances: ", diffvar1))

#Determine the experimental mean
mns <- NULL
for (i in 1 : runs){
  mns = append(mns, mean(rexp(n, lambda)))
  }

#Calculate theoretical values for the average and the variance
theomean2 <- theomean1
theovar2 <- (1/(lambda^2))/n
print(paste("Theoretical mean: ", theomean2, "Theoretical variance: ", theovar2))

#Create a an normal approximation
binwidth <- 0.5 
x <- seq(-4, 4, length=100)*theovar2 + theomean2
hx <- binwidth*runs*dnorm(x, mean=theomean2, sd=sqrt(theovar2))

#Plot the histogram of the experiments and the normal approximation
hist(mns, breaks=seq(1,9,by=binwidth),main="Histogram of the average samples",
     xlab="Sample average")
lines(x, hx)

#Create the quartile values for the experimental and theoretical values
histmns <- hist(mns, breaks=seq(1,9,by=binwidth), plot=FALSE)
fmns<- histmns$density
pmns <- pnorm(fmns, mean=theomean2, sd=sqrt(theovar2))
qmns <- qnorm(pmns, mean=theomean2, sd=sqrt(theovar2))

px <- pnorm(hx/(binwidth*runs), mean=theomean2, sd=sqrt(theovar2))
qx <- qnorm(px, mean=theomean2, sd=sqrt(theovar2))

#plot the theorethical and experimental values in a single plot
plot(qx, qx, main="Normal approximation of experimental data", xlab="quartile",
     ylab="quartile")
lines(qmns, qmns)

```