---
title: "Assignment 1 basic inferential analysis"
author: "Joris Schut"
date: "Tuesday, March 03, 2015"
output: html_document
---

##Introduction
In this basic inferential analysis differnt levels in the supp and dose variables from the ToothGrowth dataset are compared to each other to see if they have similar means. This is done by using t-test. The process of loading the data, preparing the data, performing the t-tests and deriving conclusions from these test is described in this document.

##1. Load the ToothGrowth data and perform some basic exploratory data analyses

First the packages used for this study are loaded into R. For this analysis the datastets and dplyr packages were used. The datasets package provides the data and dplyr provided extra selection and filter functionality. Second, the dataset is stored in the data variable. This variable is of class tbl_df (part of dplyr). Third, some basic exploratory data analysis is performed and some basic information about the data is printed.

```{r, echo=FALSE, message=FALSE}
#load libraries
library(datasets)
library(dplyr)

#Read the data
data<-tbl_df(ToothGrowth)

#Exploratory data analysis
print("Variable names: ")
print(names(data))
print("Dimensions of the data set (rows columns): ")
print(dim(data))
print("Class of the len variable: ")
print(class(data$len))
print("Class of the supp variable: ")
print(class(data$supp))
print("Class of the dose variable: ")
print(class(data$dose))

print("Values of the supp variable: ")
print(unique(data$supp))
print("Values of the dose variable: ")
print(unique(data$dose))
```

##2. Provide a basic summary of the data

Summary data is provided by using the summary() function.
```{r, echo=FALSE}
summary (data)
```

##3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)

T-tests are used to determine if there is a statistacally significant difference between the different groups with in the supp variable (OJ & VC) and dose variable (0.5, 1.0 & 2.0) exists.

First, we will look at the supp variable. To prepare for the t-test the differnt levels of this variable are filtered by. Then the len column is selected and the variables are bound together by colomn.
```{r, echo=FALSE}
#Create 2 data sets that contain the len entries filtered by the values of the sub variable
OJ <- filter(data, supp=="OJ") %>%
      select(len)
VC <- filter(data, supp=="VC") %>%
      select(len)

#Combine the two in a single variable
x1 <- cbind(OJ, VC)
```

After this is done a 2-sided t-test is carried out using the following hypotheses:
- H0: Xbar1 equals Xbar2
- Ha: Xbar1 does not equals Xbar2

With X1 being the data related to OJ values in the supp variable and X2 being the data related to the VC values in the supp variable.

```{r}, echo=FALSE}
#Perform a two-sided t-test
t.test(x1[,1], x1[,2], alternative="two.sided")
```

Second, we will look at the dose variable. To prepare for the t-test the differnt levels of this variable are filtered by. Then the len column is selected and the variables are bound together by colomn.

```{r, echo=FALSE}
#Create 3 data sets that contain the len entries filtered by the values of the dose variable
dose0.5 <- filter(data, dose==0.5)%>%
            select(len)
dose1.0 <- filter(data, dose==1.0)%>%
            select(len)
dose2.0 <- filter(data, dose==2.0)%>%
            select(len)

#Combine the three in a single variable
x2 <- cbind(dose0.5, dose1.0, dose2.0)
```

After this is done three 2-sided t-test are carried out using the following hypotheses.
With X1 being the data related to 0.5 values in the dose variable, X2 being the data related to 1.0 values in the dose variable and X3 being the data related to 2.0 values in the dose variable.

dose0.5 and dose1.0:
H0: Xbar1 equals Xbar2
Ha: Xbar1 does not equals Xbar2
```{r, echo=FALSE}
#Perform a two-sided t-test
t.test(x2[,1], x2[,2], alternative="two.sided")
```

dose0.5 and dose2.0:
H0: Xbar1 equals Xbar3
Ha: Xbar1 does not equals Xbar3
```{r, echo=FALSE}
#Perform a two-sided t-test
t.test(x2[,1], x2[,3], alternative="two.sided")
```

dose1.0 and dose2.0
H0: Xbar2 equals Xbar3
Ha: Xbar2 does not equals Xbar3
```{r, echo=FALSE}
#Perform a two-sided t-test
t.test(x2[,2], x2[,3], alternative="two.sided")
```

##4 State your conclusions and the assumptions needed for your conclusions

With regard to the questions wherer the means of the observations of the 2 levels present in the subb variable are similar, it can be concluded that we accept H0 in favor of Ha as the p-value of the t-test is larger than 0.05 (p = 0.06063). For the test preformed on the different values of the dose variable however, we reject H0 as these p-values are smaller than 0.05 (p-values: 1.268e-07, 4.398e-14 and 1.906e-05).

Four assumptions have been made to arrive at these conlusions: the assumption the sample are indepent and identical distributed, the samples are taken from comparable groups, the samples are not paired and the variances were unequal. Given the data is from the same study, the first two assumptions can be considered reasonable. This is also true for the third assumption as data represent differnt persons. The fourth assumption was not tested but can easily be verified (see annex).

\newpage

##Annex
Code used in this analysis:
```{r, results='hide', message=FALSE}
#load libraries
library(datasets)
library(dplyr)

#Read the data
data<-tbl_df(ToothGrowth)

#Exploratory data analysis
print("Variable names: ")
print(names(data))
print("Dimensions of the data set (rows columns): ")
print(dim(data))
print("Class of the len variable: ")
print(class(data$len))
print("Class of the supp variable: ")
print(class(data$supp))
print("Class of the dose variable: ")
print(class(data$dose))

print("Values of the supp variable: ")
print(unique(data$supp))
print("Values of the dose variable: ")
print(unique(data$dose))

summary (data)

#Create 2 data sets that contain the len entries filtered by the values of the sub variable
OJ <- filter(data, supp=="OJ") %>%
      select(len)
VC <- filter(data, supp=="VC") %>%
      select(len)

#Combine the two in a single variable
x1 <- cbind(OJ, VC)

#Perform a two-sided t-test (sup variable)
t.test(x1[,1], x1[,2], alternative="two.sided")

#Create 3 data sets that contain the len entries filtered by the values of the dose variable
dose0.5 <- filter(data, dose==0.5)%>%
            select(len)
dose1.0 <- filter(data, dose==1.0)%>%
            select(len)
dose2.0 <- filter(data, dose==2.0)%>%
            select(len)

#Combine the three in a single variable
x2 <- cbind(dose0.5, dose1.0, dose2.0)
#Perform a two-sided t-test (dose variable)
t.test(x2[,1], x2[,2], alternative="two.sided")
t.test(x2[,1], x2[,3], alternative="two.sided")
t.test(x2[,2], x2[,3], alternative="two.sided")

#Test if variances are equal
print(var(x1[,1])==var(x1[,2]))
print(var(x2[,1])==var(x2[,3]))
print(var(x2[,2])==var(x2[,3]))
```